# Functional Analysis in R

##### Loading in Data and Libraries

```{r Load data and libraries}

library(tidyverse)
library(phyloseq)
library(ggpicrust2)
library(DESeq2)
library(pheatmap)
library(ggpubr)

#### Load data 

# Metadata - extractihg it from the phyloseq object
meta = readRDS('phyloseq_taxonomy.rds') %>% .@sam_data %>% data.frame() %>% 
  rownames_to_column('sample_name')

# KEGG orthology data (PICRUSt2 output)
ko = read.delim('pred_metagenome_unstrat.tsv', row.names = 1)

# Filtering KO terms to only include PD patient samples
ko <- ko %>%
  select(all_of(meta$sample_name))

# Metacyc data (PICRUSt2 output)
metacyc = read.delim('path_abun_unstrat.tsv', row.names = 1)

# Filtering metacyc pathways to only include PD patient samples
metacyc <- metacyc %>%
  select(all_of(meta$sample_name))
```

##### Assigning Function - Pathway Error Bar Function

```{r assigning function}

# Modifying pathway errorbar function to increase space between bar plot and text
pathway_errorbar_fixed <- function (abundance, daa_results_df, Group, ko_to_kegg = FALSE, 
                            wrap_label = F, wraplength=20, fc_cutoff = 0, order_by_log = F,
                            p_values_threshold = 0.05, order = "group", select = NULL, 
                            p_value_bar = TRUE, colors = NULL, x_lab = NULL){
    require(stringr)
    errorbar_abundance_mat <- as.matrix(abundance)
    pseudo = min(abundance[abundance > 0])/2
    relative_abundance_mat <- apply(t(errorbar_abundance_mat), 
        1, function(x) (x + pseudo)/sum(x))
    sub_relative_abundance_mat <- relative_abundance_mat[rownames(relative_abundance_mat) %in% 
        (daa_results_df %>% filter(!is.na(p_adjust), p_adjust < 
            p_values_threshold) %>% pull(feature)), , drop = F]
    error_bar_matrix <- cbind(sample = colnames(sub_relative_abundance_mat), 
        group = Group, as.data.frame(t(sub_relative_abundance_mat)))
    error_bar_df <- as.data.frame(error_bar_matrix)
    error_bar_df$group <- factor(Group, levels = levels(as.factor(Group)))
    error_bar_pivot_longer_df <- tidyr::pivot_longer(error_bar_df, 
        -c(sample, group)) %>% as_tibble()
    error_bar_pivot_longer_tibble_summarised <- error_bar_pivot_longer_df %>% 
        group_by(name, group) %>% summarise(mean = mean(value)) %>% 
        ungroup() %>% group_by(name) %>% arrange(group) %>% summarise(log2FC = log2(mean[1L]/mean[2L]))
    passes_fc = error_bar_pivot_longer_tibble_summarised %>% 
        filter(abs(log2FC) > fc_cutoff) %>% pull(name)
    log_order = error_bar_pivot_longer_tibble_summarised %>% 
        arrange(log2FC) %>% pull(name)
    missing_pathways <- daa_results_df[is.na(daa_results_df$pathway_name), 
        "feature"]
    if (length(missing_pathways) > 0) {
        message("The following pathways are missing annotations and have been excluded: ", 
            paste(missing_pathways, collapse = ", "))
        message("You can use the 'pathway_annotation' function to add annotations for these pathways.")
    }
    column_names <- colnames(daa_results_df)
    group_columns <- grepl("^group", column_names)
    if (sum(group_columns) > 2) {
        p_value_bar <- FALSE
        message("There are more than two 'group' columns in the 'daa_results_df' data frame. As a result, it is not possible to compute the log2 fold values. The 'p_value_bar' has been automatically set to FALSE.")
    }
    daa_results_df <- daa_results_df[!is.na(daa_results_df[, 
        x_lab]), ]
    if (is.null(x_lab)) {
        if (ko_to_kegg == TRUE) {
            x_lab <- "pathway_name"
        }
        else {
            x_lab <- "description"
        }
        if (is.null(daa_results_df$pathway_name) & is.null(daa_results_df$description)) {
            message("Please utilize the 'pathway_annotation' function to annotate the 'daa_results_df' data frame.")
        }
    }
    if (!(x_lab %in% colnames(daa_results_df))) {
        message("The 'x_lab' you defined does not exist as a column in the 'daa_results_df' data frame.")
    }
    if (nlevels(factor(daa_results_df$method)) != 1) {
        message("The 'method' column in the 'daa_results_df' data frame contains more than one method. Please filter it to contain only one method.")
    }
    if (nlevels(factor(daa_results_df$group1)) != 1 || nlevels(factor(daa_results_df$group2)) != 
        1) {
        message("The 'group1' or 'group2' column in the 'daa_results_df' data frame contains more than one group. Please filter each to contain only one group.")
    }
    if (is.null(colors)) {
        colors <- c("#d93c3e", "#3685bc", "#6faa3e", "#e8a825", 
            "#c973e6", "#ee6b3d", "#2db0a7", "#f25292")[1:nlevels(as.factor(Group))]
    }
    errorbar_abundance_mat <- as.matrix(abundance)
    daa_results_filtered_df <- daa_results_df[daa_results_df$p_adjust < 
        p_values_threshold & daa_results_df$feature %in% passes_fc, 
        ]
    if (!is.null(select)) {
        daa_results_filtered_sub_df <- daa_results_filtered_df[daa_results_filtered_df$feature %in% 
            select, ]
    }
    else {
        daa_results_filtered_sub_df <- daa_results_filtered_df
    }
    if (nrow(daa_results_filtered_sub_df) > 30) {
        message(paste0("The number of features with statistical significance exceeds 30, leading to suboptimal visualization. ", 
            "Please use 'select' to reduce the number of features.\n", 
            "Currently, you have these features: ", paste(paste0("\"", 
                daa_results_filtered_sub_df$feature, "\""), collapse = ", "), 
            ".\n", "You can find the statistically significant features with the following command:\n", 
            "daa_results_df %>% filter(p_adjust < 0.05) %>% select(c(\"feature\",\"p_adjust\"))"))
        stop()
    }
    if (nrow(daa_results_filtered_sub_df) == 0) {
        stop("Visualization with 'pathway_errorbar' cannot be performed because there are no features with statistical significance. ", 
            "For possible solutions, please check the FAQ section of the tutorial.")
    }
    relative_abundance_mat <- apply(t(errorbar_abundance_mat), 
        1, function(x) x/sum(x))
    sub_relative_abundance_mat <- relative_abundance_mat[rownames(relative_abundance_mat) %in% 
        daa_results_filtered_sub_df$feature, , drop = F]
    error_bar_matrix <- cbind(sample = colnames(sub_relative_abundance_mat), 
        group = Group, t(sub_relative_abundance_mat))
    error_bar_df <- as.data.frame(error_bar_matrix)
    error_bar_df$group <- factor(Group, levels = levels(as.factor(Group)))
    error_bar_pivot_longer_df <- tidyr::pivot_longer(error_bar_df, 
        -c(sample, group))
    error_bar_pivot_longer_tibble <- mutate(error_bar_pivot_longer_df, 
        group = as.factor(group))
    error_bar_pivot_longer_tibble$sample <- factor(error_bar_pivot_longer_tibble$sample)
    error_bar_pivot_longer_tibble$name <- factor(error_bar_pivot_longer_tibble$name)
    error_bar_pivot_longer_tibble$value <- as.numeric(error_bar_pivot_longer_tibble$value)
    error_bar_pivot_longer_tibble_summarised <- error_bar_pivot_longer_tibble %>% 
        group_by(name, group) %>% summarise(mean = mean(value), 
        sd = stats::sd(value))
    error_bar_pivot_longer_tibble_summarised <- error_bar_pivot_longer_tibble_summarised %>% 
        mutate(group2 = "nonsense")
    switch(order, p_values = {
        order <- order(daa_results_filtered_sub_df$p_adjust)
    }, name = {
        order <- order(daa_results_filtered_sub_df$feature)
    }, group = {
        daa_results_filtered_sub_df$pro <- 1
        for (i in levels(error_bar_pivot_longer_tibble_summarised$name)) {
            error_bar_pivot_longer_tibble_summarised_sub <- error_bar_pivot_longer_tibble_summarised[error_bar_pivot_longer_tibble_summarised$name == 
                i, ]
            pro_group <- error_bar_pivot_longer_tibble_summarised_sub[error_bar_pivot_longer_tibble_summarised_sub$mean == 
                max(error_bar_pivot_longer_tibble_summarised_sub$mean), 
                ]$group
            pro_group <- as.vector(pro_group)
            daa_results_filtered_sub_df[daa_results_filtered_sub_df$feature == 
                i, ]$pro <- pro_group
        }
        order <- order(daa_results_filtered_sub_df$pro, daa_results_filtered_sub_df$p_adjust)
    }, pathway_class = {
        if (!"pathway_class" %in% colnames(daa_results_filtered_sub_df)) {
            stop("The 'pathway_class' column is missing in the 'daa_results_filtered_sub_df' data frame. ", 
                "Please use the 'pathway_annotation' function to annotate the 'pathway_daa' results.")
        }
        order <- order(daa_results_filtered_sub_df$pathway_class, 
            daa_results_filtered_sub_df$p_adjust)
    }, {
        order <- order
    })
    daa_results_filtered_sub_df <- daa_results_filtered_sub_df[order, 
        ]
    error_bar_pivot_longer_tibble_summarised_ordered <- data.frame(name = NULL, 
        group = NULL, mean = NULL, sd = NULL)
    for (i in daa_results_filtered_sub_df$feature) {
        error_bar_pivot_longer_tibble_summarised_ordered <- rbind(error_bar_pivot_longer_tibble_summarised_ordered, 
            error_bar_pivot_longer_tibble_summarised[error_bar_pivot_longer_tibble_summarised$name == 
                i, ])
    }
    if (ko_to_kegg == FALSE) {
        error_bar_pivot_longer_tibble_summarised_ordered[, x_lab] <- rep(daa_results_filtered_sub_df[, 
            x_lab], each = length(levels(factor(error_bar_pivot_longer_tibble_summarised_ordered$group))))
    }
    if (ko_to_kegg == TRUE) {
        error_bar_pivot_longer_tibble_summarised_ordered$pathway_class <- rep(daa_results_filtered_sub_df$pathway_class, 
            each = length(levels(factor(error_bar_pivot_longer_tibble_summarised_ordered$group))))
    }
    if (order_by_log == F) {
        error_bar_pivot_longer_tibble_summarised_ordered$name <- factor(error_bar_pivot_longer_tibble_summarised_ordered$name, 
            levels = rev(daa_results_filtered_sub_df$feature))
    }
    else {
        error_bar_pivot_longer_tibble_summarised_ordered$name <- factor(error_bar_pivot_longer_tibble_summarised_ordered$name, 
            levels = rev(log_order))
    }
    if ("pathway_name" %in% names(daa_results_filtered_sub_df)) {
        daa_results_filtered_sub_df$pathway_name = sapply(daa_results_filtered_sub_df$pathway_name, 
            function(x) str_split(x, " \\[")[[1]][1] %>% str_trim())
        if (wrap_label == T) {
            daa_results_filtered_sub_df$pathway_name = stringr::str_wrap(daa_results_filtered_sub_df$pathway_name, 
                wraplength)
        }
    }
    bar_errorbar <- ggplot2::ggplot(error_bar_pivot_longer_tibble_summarised_ordered, 
        ggplot2::aes(mean, name, fill = group)) + ggplot2::geom_errorbar(ggplot2::aes(xmax = mean + 
        sd, xmin = 0), position = ggplot2::position_dodge(width = 0.8), 
        width = 0.5, size = 0.5, color = "black") + ggplot2::geom_bar(stat = "identity", 
        position = ggplot2::position_dodge(width = 0.8), width = 0.8) + 
        GGally::geom_stripped_cols(width = 10) + ggplot2::scale_fill_manual(values = colors) + 
        ggplot2::scale_color_manual(values = colors) + ggprism::theme_prism() + 
        ggplot2::scale_x_continuous(expand = c(0, 0)) + ggplot2::scale_y_discrete(labels = rev(daa_results_filtered_sub_df[, 
        x_lab])) + ggplot2::labs(x = "Relative Abundance", y = NULL) + 
        ggplot2::theme(axis.ticks.y = ggplot2::element_blank(), 
            axis.line.y = ggplot2::element_blank(), axis.line.x = ggplot2::element_line(size = 0.5), 
            axis.ticks.x = ggplot2::element_line(size = 0.5), 
            panel.grid.major.y = ggplot2::element_blank(), panel.grid.major.x = ggplot2::element_blank(), 
            axis.text = ggplot2::element_text(size = 13, color = "black"), 
            axis.text.x = ggplot2::element_text(margin = ggplot2::margin(r = 0)), 
            axis.text.y = ggplot2::element_text(size = 13, color = "black", 
                margin = ggplot2::margin(b = 6, r = 10)), axis.title.x = ggplot2::element_text(size = 13, 
                color = "black", hjust = 0.5), legend.position = "top", 
            legend.key.size = ggplot2::unit(0.3, "cm"), legend.direction = "vertical", 
            legend.justification = "left", legend.text = ggplot2::element_text(size = 13, 
                face = "bold"), legend.box.just = "right", plot.margin = ggplot2::margin(0, 
                0.5, 0.5, 3, unit = "cm")) + ggplot2::coord_cartesian(clip = "off")
    if (ko_to_kegg == TRUE) {
        pathway_class_group_mat <- daa_results_filtered_sub_df$pathway_class %>% 
            table() %>% data.frame() %>% column_to_rownames(".")
        pathway_class_group <- data.frame(. = unique(daa_results_filtered_sub_df$pathway_class), 
            Freq = pathway_class_group_mat[unique(daa_results_filtered_sub_df$pathway_class), 
                ])
        start <- c(1, rev(pathway_class_group$Freq)[1:(length(pathway_class_group$Freq) - 
            1)]) %>% cumsum()
        end <- cumsum(rev(pathway_class_group$Freq))
        ymin <- start - 1/2
        ymax <- end + 1/2
        nPoints <- length(start)
        pCol <- c("#D51F26", "#272E6A", "#208A42", "#89288F", 
            "#F47D2B", "#FEE500", "#8A9FD1", "#C06CAB", "#E6C2DC", 
            "#90D5E4", "#89C75F", "#F37B7D", "#9983BD", "#D24B27", 
            "#3BBCA8", "#6E4B9E", "#0C727C", "#7E1416", "#D8A767", 
            "#3D3D3D")[1:nPoints]
        pFill <- pCol
        for (i in 1:nPoints) {
            bar_errorbar <- bar_errorbar + ggplot2::annotation_custom(grob = grid::rectGrob(gp = grid::gpar(col = pCol[i], 
                fill = pFill[i], lty = NULL, lwd = NULL, alpha = 0.2)),
                xmin = -2, xmax = 0,
                ymin = ymin[i], ymax = ymax[i])
        }
    }
    daa_results_filtered_sub_df <- cbind(daa_results_filtered_sub_df, 
        negative_log10_p = -log10(daa_results_filtered_sub_df$p_adjust), 
        group_nonsense = "nonsense", log_2_fold_change = NA)
    for (i in daa_results_filtered_sub_df$feature) {
        mean <- error_bar_pivot_longer_tibble_summarised_ordered[error_bar_pivot_longer_tibble_summarised_ordered$name %in% 
            i, ]$mean
        daa_results_filtered_sub_df[daa_results_filtered_sub_df$feature == 
            i, ]$log_2_fold_change <- log2(mean[1]/mean[2])
    }
    if (order_by_log == F) {
        daa_results_filtered_sub_df$feature <- factor(daa_results_filtered_sub_df$feature, 
            levels = rev(daa_results_filtered_sub_df$feature))
    }
    else {
        daa_results_filtered_sub_df$feature <- factor(daa_results_filtered_sub_df$feature, 
            levels = rev(log_order))
    }
    p_values_bar <- daa_results_filtered_sub_df %>% ggplot2::ggplot(ggplot2::aes(feature, 
        log_2_fold_change, fill = group_nonsense)) + ggplot2::geom_bar(stat = "identity", 
        position = ggplot2::position_dodge(width = 0.8), width = 0.8) + 
        ggplot2::labs(y = "Log2 Fold Change", x = NULL) + GGally::geom_stripped_cols() + 
        ggplot2::scale_fill_manual(values = "#87ceeb") + ggplot2::scale_color_manual(values = "#87ceeb") + 
        ggplot2::geom_hline(ggplot2::aes(yintercept = 0), linetype = "dashed", 
            color = "black") + ggprism::theme_prism() + ggplot2::theme(axis.ticks.y = ggplot2::element_blank(), 
        axis.line.y = ggplot2::element_blank(), axis.line.x = ggplot2::element_line(size = 0.5), 
        axis.ticks.x = ggplot2::element_line(size = 0.5), panel.grid.major.y = ggplot2::element_blank(), 
        panel.grid.major.x = ggplot2::element_blank(), axis.text = ggplot2::element_text(size = 13, 
            color = "black"), axis.text.y = ggplot2::element_blank(), 
        axis.text.x = ggplot2::element_text(size = 13, color = "black", 
            margin = ggplot2::margin(b = 6)), axis.title.x = ggplot2::element_text(size = 14, 
            color = "black", hjust = 0.5), legend.position = "non") + 
        ggplot2::coord_flip()
    if (ko_to_kegg == TRUE) {
        pathway_class_y <- (ymax + ymin)/2 - 0.5
        pathway_class_plot_df <- as.data.frame(cbind(nonsense = "nonsense", 
            pathway_class_y = pathway_class_y, pathway_class = rev(unique(daa_results_filtered_sub_df$pathway_class))))
        pathway_class_plot_df$pathway_class = sapply(pathway_class_plot_df$pathway_class, 
            function(x) str_split(x, ";")[[1]] %>% .[length(.)] %>% 
                str_trim())
        pathway_class_plot_df$pathway_class = sapply(pathway_class_plot_df$pathway_class, 
            function(x) str_remove_all(x, "KEGG Orthology [(]KO[)] \\[BR:ko00001\\];  "))
        pathway_class_plot_df$pathway_class_y <- as.numeric(pathway_class_plot_df$pathway_class_y)
        if (nrow(pathway_class_plot_df) == 2) {
            pathway_class_plot_df$pathway_class_y = 1
            pathway_class_plot_df = pathway_class_plot_df[1, 
                , drop = F]
        }
        pathway_class_annotation <- pathway_class_plot_df %>% 
            ggplot2::ggplot(ggplot2::aes(nonsense, pathway_class_y)) + 
            ggplot2::geom_text(ggplot2::aes(nonsense, pathway_class_y, 
                label = pathway_class), size = 4.5, color = "black", 
                fontface = "bold", family = "sans") + ggplot2::scale_y_discrete(position = "right") + 
            ggprism::theme_prism() + ggplot2::theme(axis.ticks = ggplot2::element_blank(), 
            axis.line = ggplot2::element_blank(), panel.grid.major.y = ggplot2::element_blank(), 
            panel.grid.major.x = ggplot2::element_blank(), panel.background = ggplot2::element_blank(), 
            axis.text = ggplot2::element_blank(), plot.margin = ggplot2::unit(c(0, 
                0.2, 0, 0), "cm"), axis.title.y = ggplot2::element_blank(), 
            axis.title.x = ggplot2::element_blank(), legend.position = "non")
    }
    daa_results_filtered_sub_df$p_adjust <- as.character(signif(daa_results_filtered_sub_df$p_adjust, 
        2))
    daa_results_filtered_sub_df$unique <- nrow(daa_results_filtered_sub_df) - 
        seq_len(nrow(daa_results_filtered_sub_df)) + 1
    p_annotation <- daa_results_filtered_sub_df %>% ggplot2::ggplot(ggplot2::aes(group_nonsense, 
        p_adjust)) + ggplot2::geom_text(ggplot2::aes(group_nonsense, 
        unique, label = p_adjust), size = 5, color = "black", 
        fontface = "bold", family = "sans") + ggplot2::labs(y = "p-value (adjusted)") + 
        ggplot2::scale_y_discrete(position = "right") + ggprism::theme_prism() + 
        ggplot2::theme(axis.ticks = ggplot2::element_blank(), 
            axis.line = ggplot2::element_blank(), panel.grid.major.y = ggplot2::element_blank(), 
            panel.grid.major.x = ggplot2::element_blank(), panel.background = ggplot2::element_blank(), 
            axis.text = ggplot2::element_blank(), plot.margin = ggplot2::unit(c(0, 
                0.2, 0, 0), "cm"), axis.title.y = ggplot2::element_text(size = 14, 
                color = "black", vjust = 0), axis.title.x = ggplot2::element_blank(), 
            legend.position = "none")
    if (p_value_bar == TRUE) {
        if (ko_to_kegg == TRUE) {
            combination_bar_plot <- pathway_class_annotation + 
                bar_errorbar + p_values_bar + p_annotation + 
                patchwork::plot_layout(ncol = 4, widths = c(1.2, 
                  1, 0.5, 0.3))
        }
        else {
            combination_bar_plot <- bar_errorbar + p_values_bar + 
                p_annotation + patchwork::plot_layout(ncol = 3, 
                widths = c(2.3, 0.7, 0.5))
        }
    }
    else {
        if (ko_to_kegg == TRUE) {
            combination_bar_plot <- pathway_class_annotation + 
                bar_errorbar + p_annotation + patchwork::plot_layout(ncol = 3, 
                widths = c(1, 1.2, 0.3))
        }
        else {
            combination_bar_plot <- bar_errorbar + p_annotation + 
                patchwork::plot_layout(ncol = 2, widths = c(2.5, 
                  0.4))
        }
    }
    return(combination_bar_plot)
}
```

## Functional Analysis with KEGG Orthology Terms

##### Deseq2 functional analysis of KO terms

```{r Deseq2 KEGG analysis}

# Perform pathway differential abundance analysis (DAA) using DESeq 
daa_results_df = pathway_daa(abundance = ko,
                             metadata = meta, 
                             group = "Lived_on_farm", 
                             daa_method = "DESeq2", 
                             select = NULL, reference = NULL)


# Annotate pathway results using KO to KEGG conversion
daa_annotated_results_df = pathway_annotation(pathway = "KO",
                                              daa_results_df = daa_results_df,
                                              ko_to_kegg = TRUE)

```

##### Generating error bar plot of functional differential abundance results

```{r pathway errorbar plot}

# Given that we had 14 hits total, we will not filter out any and will keep all our results
peb = pathway_errorbar_fixed(abundance = ko, 
                     daa_results_df = daa_annotated_results_df, 
                     Group = meta$Lived_on_farm, 
                     wrap_label = T, wraplength=100,
                     fc_cutoff = F, order_by_log = F,
                     p_values_threshold = 0.05, 
                     order = "pathway_class", 
                     ko_to_kegg = TRUE, 
                     p_value_bar = TRUE, 
                     x_lab = "pathway_name")
peb


# This plot has an unclassified pathway, which must be filtered out for the manuscript-ready plot
daa_annotated_results_df <- daa_annotated_results_df %>% 
  filter(feature != "ko:K09963")
  
# Now plot without the unknown pathway
peb = pathway_errorbar_fixed(abundance = ko, 
                     daa_results_df = daa_annotated_results_df, 
                     Group = meta$Lived_on_farm, 
                     wrap_label = T, wraplength=100,
                     fc_cutoff = F, order_by_log = F,
                     p_values_threshold = 0.05, 
                     order = "pathway_class", 
                     ko_to_kegg = TRUE, 
                     p_value_bar = TRUE, 
                     x_lab = "pathway_name")
peb



# ggsave("KEGG_Error_Bar_Fixed_DESeq2.png",
#        plot = peb, width = 22, height = 10)
```

##### Making heat map to evaluate which features are highly correlated (essentially the same)

```{r identical features}

# Filtering for p value and log2 fold change
sig_features = daa_annotated_results_df %>% 
  filter(p_adjust <= 0.05, abs(log2_fold_change) > 0.5) %>% 
  pull('feature') %>% unique()

ko_relab = read.delim('pred_metagenome_unstrat.tsv',row.names = 1) %>% 
  apply(2,function(x) x/sum(x)) %>% as.data.frame()
colSums(ko_relab) # Should be all 1

# Spearman tests between every feature pair
stats = ko_relab %>% t() %>%  
  as.data.frame() %>% 
  select(all_of(sig_features)) %>% 
  cor(method = 'spearman') 

# Define the color palette
color_palette <- colorRampPalette(c("blue", "white", "red"))(40)

# Define breaks for the color scale
breaks <- seq(-1, 1, length.out = 41)  # 40 colors + 1 for the endpoint

# Create the clustered heatmap with centered color at zero
pheatmap(stats, 
         clustering_distance_rows = "euclidean",  
         clustering_distance_cols = "euclidean", 
         clustering_method = "complete",          
         color = color_palette, 
         breaks = breaks,
         main = '',
         fontsize_row = 10, 
         fontsize_col = 10,
         filename = "heatmap_deseq2.png",
         height= 9, width = 9)





# Based on heat map, clustering by KO clusters into the 2 visibly distinct clusters
# Cluster red = Cluster A
cluster_red <- c(
  "ko:K17236",
  "ko:K04076",
  "ko:K05822",
  "ko:K20447",
  "ko:K21900",
  "ko:K08153",
  "ko:K15519"
)

# Cluster blue = Cluster B
cluster_blue <- c(
  "ko:K00023",
  "ko:K05795",
  "ko:K18650",
  "ko:K00561",
  "ko:K05881",
  "ko:K07455"
)

ko_red <- ko[rownames(ko) %in% cluster_red, ]
ko_blue <- ko[rownames(ko) %in% cluster_blue, ]

nrow(ko_red)   # Sanity check - this column has 7 rows which is expected
nrow(ko_blue)  # Sanity check - this column has 6 rows which is expected

red_sum <- colSums(ko_red)
blue_sum <- colSums(ko_blue)


#### Cluster Red (A)
# Making a data frame for cluster
red_df <- data.frame(
  Sample = colnames(ko_red),
  Abundance = log2(red_sum + 1),
  farm_living = meta$Lived_on_farm
)

# Plotting cluster A (red) KO terms farm-living vs non farm-living
red_plot <- ggplot(red_df, aes(x = farm_living, y = Abundance, fill = farm_living)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.2) +
  ylab("Log2 Cluster A Abundance") +
  xlab("Lived on a Farm") +
  scale_fill_manual(values = c("No" = "darkgrey", "Yes" = "steelblue"), name = "Lived on a Farm") +
  theme_classic(base_size = 50) +
  theme(legend.position = "none") +
  stat_compare_means(comparisons = list(c("Yes", "No")), method = "wilcox.test", size = 18)

red_plot


#### Cluster Blue (B)
# Making a data frame for cluster
blue_df <- data.frame(
  Sample = colnames(ko_blue),
  Abundance = log2(blue_sum + 1),
  farm_living = meta$Lived_on_farm
)

# Plotting cluster B (blue) KO terms farm-living vs non farm-living
blue_plot <- ggplot(blue_df, aes(x = farm_living, y = Abundance, fill = farm_living)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.2) +
  ylab("Log2 Cluster B Abundance") +
  xlab("Lived on a Farm") +
  scale_fill_manual(values = c("No" = "darkgrey", "Yes" = "steelblue"), name = "Lived on a Farm") +
  theme_classic(base_size = 50) +
  theme(legend.position = "none") +
  stat_compare_means(comparisons = list(c("Yes", "No")), method = "wilcox.test", size = 18)

blue_plot


# ggsave("ClusterA_bar_plot2.png",
#        plot = red_plot, width = 20, height = 18, dpi = 300)
# ggsave("ClusterB_bar_plot2.png",
#        plot = blue_plot, width = 20, height = 18, dpi = 300)




## Evaluating whether Cluster A and CLuster B are correlated
cor.test(red_sum, blue_sum, method = "spearman")

```

##### PCA Plot to see how similar farm-living vs non-farm-living groups are

```{r PCA}

# Generate pathway PCA plot
pca_plot <- pathway_pca(abundance = ko,
            metadata = meta, 
            group = "Lived_on_farm")

pca_plot

# Saving the PCA plot 
# ggsave("pca_deseq2.png",
#        plot = pca_plot, width = 20, height = 14)
```

## Functional Analysis with Metacyc Pathways

##### Deseq2 functional analysis of metacyc pathways

```{r Deseq2 Metacyc analysis}

# Perform pathway differential abundance analysis (DAA) using DESeq with metacyc data
daa_results_metacyc_df = pathway_daa(abundance = metacyc,
                             metadata = meta, 
                             group = "Lived_on_farm", 
                             daa_method = "DESeq2", 
                             select = NULL, reference = NULL)


# Annotate the results (similar to what was done above with the KO terms, but instead using Metacyc pathways)
daa_annotated_results_metacyc_df = pathway_annotation(pathway = "MetaCyc",
                                              daa_results_df = daa_results_metacyc_df,
                                              ko_to_kegg = FALSE)
```

##### Generating error bar plot of functional differential abundance results

```{r pathway error bar plot}

# Filtering for adjusted p-value <= 0.05 
daa_annotated_results_metacyc_df <- daa_annotated_results_metacyc_df %>% filter(p_adjust <= 0.05)

# Generating error bar plot
peb_metacyc = pathway_errorbar_fixed(abundance = metacyc, 
                     daa_results_df = daa_annotated_results_metacyc_df,
                     Group = meta$Lived_on_farm, 
                     wrap_label = T, wraplength=100,
                     fc_cutoff = F, order_by_log = F,
                     p_values_threshold = 1.0, 
                     order = "group", 
                     ko_to_kegg = FALSE, 
                     p_value_bar = TRUE, 
                     x_lab = "feature")
peb_metacyc


# ggsave("Metacyc_Error_Bar_Fixed_DESeq2_2.png",
#        plot = peb_metacyc, width = 22, height = 10)
```

##### Making heat map to evaluate which features are highly correlated (essentially the same)

```{r metacyc heat map}

# Filtering adjusted p-value 
metacyc_sig_features = daa_annotated_results_metacyc_df %>% 
  filter(p_adjust <= 0.05, abs(log2_fold_change) > 0) %>% 
  pull('feature') %>% unique()

metacyc_relab = read.delim('path_abun_unstrat.tsv',row.names = 1) %>% 
  apply(2,function(x) x/sum(x)) %>% as.data.frame()
colSums(metacyc_relab) # Should all be 1

# Spearman test between every feature pair
stats = metacyc_relab %>% t() %>%  
  as.data.frame() %>% 
  select(all_of(metacyc_sig_features)) %>% 
  cor(method = 'spearman') 


# Cannot create heatmap, as there is only one significant metacyc pathway
```

##### PCA Plot to see how similar farm-living vs non-farm-living groups are

```{r metacyc pca plot}

# Generate pathway PCA plot
pca_metacyc_plot <- pathway_pca(abundance = metacyc,
            metadata = meta, 
            group = "Lived_on_farm")

pca_metacyc_plot

# Saving metacyc pathways PCA plot 
# ggsave("pca_deseq2_metacyc.png",
#        plot = pca_metacyc_plot, width = 20, height = 14)
```
