---
title: "Random Forest"
author: "Lucy Perkins"
date: "2026-03-21"
output: html_document
---
# Part 1: Load Libraries and Data

```{r Load Libraries and Data}
# Load necessary libraries
library(randomForest)
library(caret)
library(ranger)
library(pROC)
library(boot)
library(ggplot2)
library(phyloseq)
library(tidyverse)

# Load the dataset
ps = readRDS('phyloseq_taxonomy.rds')
```

# Part 2: Prep Data for Random Forest

```{r Prep Data}
# Aggregate reads to the Genus level
ps <- tax_glom(ps, "Genus")

# CLR transform
ps_clr <- microbiome::transform(ps, "clr")

# Melt to long format
df <- psmelt(ps_clr)

# Create onset groups
df <- df %>%
  mutate(
    Onset_group = case_when(
      Age_onset >= 60 ~ "Later_Onset",
      Age_onset < 60 ~ "Earlier_Onset")
    )

# Remove missing values for key variables
df <- df %>%
  filter(!is.na(Onset_group), !is.na(Genus))

# Z-transform abundance within each genus
df_transformed <- df %>%
  group_by(Genus) %>%
  mutate(Abundance = as.numeric(scale(Abundance))) %>%
  ungroup()

# Summarize so each Sample x Genus has only one value
df_summarized <- df_transformed %>%
  group_by(Sample, Onset_group, Genus) %>%
  summarise(Abundance = mean(Abundance), .groups = "drop")

# Pivot to wide format
df_pivot <- df_summarized %>%
  pivot_wider(names_from = Genus, values_from = Abundance)

# Final dataframe
df_final <- df_pivot %>%
  select(-Sample) %>%
  mutate(Onset_group = factor(Onset_group))

```

# Part 3: Set up RF

Start by separating the data into predictors (taxa) and outcome (onset group)

```{r predictors_and_outcome}
# Set predictors as taxa
predictors <- df_final %>% select(-Onset_group)

# Set outcome as onset group factor
outcome <- df_final %>% 
  pull(Onset_group) %>% 
  factor(levels = c("Earlier_Onset", "Later_Onset"))
```

Next, set up the k-fold cross-validation with k = 10

```{r cross-validation}
# Randomly subset the rows into k equal bins
k <- 10
set.seed(421)
folds <- createFolds(outcome, k = k, list = TRUE)

# Each of these folds will take a turn being the test dataset
str(folds)

```

Finally, set up hyperparameter tuning

```{r hyperparamater tuning}
# Set parameters for tuning
tune_grid <- expand.grid(
  mtry = c(5, 10, 20, 30),
  splitrule = "gini",
  min.node.size = c(5, 10, 20, 50))

```

# Part 4: Run RF

Use given run_rf function using source command and examine the output

```{r run_rf}
source('randomforest_functions.R')

pd_model = run_rf(X = predictors, y = outcome, 
                  fold_list = folds,
                  hyper = tune_grid, 
                  rngseed = 421)

names(pd_model)
```

# Part 5: Interpret Results

First, generate ROC curve to interpret results

```{r ROC curve}
# Calculate ROC curve points from RF output
roc_test = roc(pd_model$test_labels$true_labels,
               pd_model$test_labels$predicted_probabilities)
roc_train = roc(pd_model$train_labels$true_labels,
                pd_model$train_labels$predicted_probabilities)

# True positive rate = sensitivity
# False positive rate = 1-specificity
plot_roc <- ggplot() +
  # Training data
  geom_line(aes(x = 1 - roc_train$specificities, 
                y = roc_train$sensitivities), 
            color = "red", linewidth = 1) +
  # Test data
  geom_line(aes(x = 1 - roc_test$specificities,
                y = roc_test$sensitivities), 
            color = "black", linewidth = 1) +
  geom_abline(slope = 1, intercept = 0, color = "gray", linetype = "dashed", linewidth = 1) +
  labs(x = "False Positive Rate", y = "True Positive Rate") +
  annotate(
    "text",
    x = 0.7,
    y = 0.2,
    label = sprintf(
      "Train (red): %.2f (%.2f-%.2f)\nTest (black): %.2f (%.2f-%.2f)",
      pd_model$auc_train, pd_model$auc_train_ci[1], pd_model$auc_train_ci[2],
      pd_model$auc_test, pd_model$auc_test_ci[1], pd_model$auc_test_ci[2]
    ),
    size = 7
  ) +
  theme_minimal(base_size = 20)

# ggsave(
#   filename = "roc_curve.png",
#   plot = plot_roc,
#   width = 8,
#   height = 6,
#   dpi = 300
# )
```
Next, compile the results into a table
```{r results table}
roc_data <- data.frame(
  Dataset = "RF Age of Onset Model",
  Training_AUC = round(pd_model$auc_train, 2),
  Training_AUC_CI = paste0(
    round(pd_model$auc_train_ci[1], 2), "-",
    round(pd_model$auc_train_ci[2], 2)
  ),
  Testing_AUC = round(pd_model$auc_test, 2),
  Testing_AUC_CI = paste0(
    round(pd_model$auc_test_ci[1], 2), "-",
    round(pd_model$auc_test_ci[2], 2)
  )
)

roc_data
```
Finally, plot the importance values of each predictor

```{r importance values}
# Plot only the top 20 most important genera
plot_importance <- pd_model$importance %>%
  slice_max(order_by = MeanDecreaseGini, n = 20) %>%
  arrange(MeanDecreaseGini) %>%
  mutate(Feature = factor(Feature, levels = Feature)) %>%
  ggplot(aes(x = Feature, y = MeanDecreaseGini, fill = MeanDecreaseGini)) +
  geom_col() +
  coord_flip() +
  theme_classic(base_size = 18) +
  theme(
    axis.text.y = element_text(size = 11),
    axis.text.x = element_text(size = 12),
    legend.position = "right",
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10)
  ) +
  ylab("Importance (Gini)") +
  xlab(NULL)

# ggsave(
#   filename = "importance_plot.png",
#   plot = plot_importance,
#   width = 10,
#   height = 7,
#   dpi = 300
# )
```

